R for Clinical Study Reports and Submission

### 2022-01-06 tlf-efficacy-km.Rmd
simplify kmplot chapter

```{r}
library(haven) # Read SAS data
library(dplyr) # Manipulate data
library(r2rtf) # Reporting in RTF format
library(survival) # Fit survival model
library(survminer) # K-M plot with risk table; optional for user's preference
library(visR) # K-M plot with risk table; optional for user's preference
library(forcats) # Reorder ordinal variable level based on preference
```

In this chapter, we illustrate how to create a Kaplan-Meier plot in a study.
For the survival analysis in efficacy, time to dermatologic event (TTDE) will be analyzed.

@@

The `survival` package is used to create the curve data points, users can use `Surv()` with formula directly, 
or use a fitted Cox model with the `coxph()` function 
if we also want to obtain other information, i.e. hazard ratio, survival rates, etc. 
<!-- TODO: "Analysis of Survival Endpoint Table" Chapter -->

@@

There are various ways to create a KM plot. In the following, we will illustrate how to use `survival`, `survminer` or `visR` to generate a KM plot with a risk table, respectively. There are pro and con for using these package. Compared with the basic `survival` package, `survminer` or `visR` package can be easily used to customize a nice-look KM plot. However, the `survminer` or `visR` package has relative heavy dependencies (Ref: https://github.com/elong0527/r4csr/issues/50#issuecomment-985109643), which may increase risk of reproducibility (Section 4.2 of the [White Paper](https://www.pharmar.org/white-paper/)).

### Plot with `survival`

Firstly, create a function `risk_tab()` to count cumulative number at risk, then display them as the risk table.

```{r}
risk_tab <- function(tlim, time) {
  n_risk <- NULL
  for (i in tlim) {
    n_risk <- c(n_risk, sum(time >= i))
  }
  return(n_risk)
}
```

Then create the KM plot and save as a `.png` file.

```{r}
# Save as a PNG file
png(
  file = "tlf/fig_km1.png",
  width = 3000,
  height = 2000,
  res = 400
)

par(mfrow = c(1, 1),
    mar = c(9, 11, 1, 1),
    family = "Times") # Set plot margins and font style

plot(
  fit,
  xlab = "Time in Months",
  ylab = "Survival probability",
  lwd = 2,
  mark.time = T,
  col = c(2, 3, 4),
  lty = c(1, 2, 3)
)
legend(
  "topright",
  lwd = 2,
  cex = 0.9,
  col = c(2, 3, 4),
  lty = c(1, 2, 3),
  legend = levels(adtte_ttde$TRTP),
  bty = "n"
)
mtext("Number at risk",
      side = 1,
      line = 4,
      at = 0)
grid <- 0:6
mtext(
  c("Placebo", risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Placebo"])),
  side = 1,
  adj = c(1, rep(NA, length(grid))),
  line = 5,
  at = c(-0.8, grid)
)
mtext(
  c("Xanomeline Low Dose",
    risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Xanomeline Low Dose"])),
  side = 1,
  adj = c(1, rep(NA, length(grid))),
  line = 6,
  at = c(-0.8, grid)
)
mtext(
  c("Xanomeline High Dose",
    risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Xanomeline High Dose"])),
  side = 1,
  adj = c(1, rep(NA, length(grid))),
  line = 7,
  at = c(-0.8, grid)
)

@@

Then we can use the `r2rtf` package to create a formatted RTF figure. 
More details can be found in <https://merck.github.io/r2rtf/articles/example-figure.html>.

```{r}
# Create RTF figure
rtf_read_png("tlf/fig_km1.png") %>% # Read the PNG file from the file path
  rtf_title(
    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
    "All Participants"
  ) %>% # Add title or subtitle
  rtf_footnote("footnote") %>% # Add footnote
  rtf_source("[datasource: adam-adtte]") %>% # Add data source
  rtf_figure(fig_width = 6, fig_height = 4) %>% # Set proportional figure size to the original PNG figure size
  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
  write_rtf(file = "tlf/fig_km1.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/fig_km1.pdf")
```

### Plot with `survminer`

Alternatively, the `survminer` package can be also used to plot the KM curve with an elegant `ggplot2` style. 
There are tons of customized options for users to create a K-M plot
with risk table and testing result (nominal p-value). 
More details can be found in `?survminer::ggsurvplot`.

```{r}
fig_km2 <- ggsurvplot(
  fit,
  xlab = "Time in Months",
  ylab = "Survival Probability",
  break.time.by = 1,
  #  xscale = "d_m",
  #  xlim = c(0, 200),
  linetype = c(1, 2, 3),
  risk.table = TRUE,
  risk.table.height = 0.3,
  # Useful when you have multiple groups
  legend = c(0.8, 0.9),
  legend.title = "",
  legend.labs = sort(unique(adtte_ttde$TRTP)),
  censor.size = 4,
  size = 0.8,
  title = "",
  censor.shape = "|",
  pval = TRUE,
  pval.method = TRUE,
  pval.coord = c(0, 0.1),
  pval.method.coord = c(0, 0.2),
  pval.size = 5,
  font.family = "Times New Roman",
  ggtheme = theme_classic2(base_family = "Times New Roman", base_size = 18)
)
```

Then we can save the figure as a `.png` file, 
and use the `r2rtf` package to create a formatted RTF figure. 
More details can be found in <https://merck.github.io/r2rtf/articles/example-figure.html>.

```{r}
# Save as PNG file
# grid.draw.ggsurvplot <- function(x) survminer:::print.ggsurvplot(x, newpage = FALSE)
ggsave("tlf/fig_km2.png",
       plot = fig_km2,
       width = 10,
       height = 6)

# Create RTF figure
rtf_read_png("tlf/fig_km2.png") %>% # Read PNG files from the file path
  rtf_title(
    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
    "All Participants"
  ) %>% # Add title or subtitle
  rtf_footnote("footnote") %>% # Add footnote
  rtf_source("[datasource: adam-adtte]") %>% # Add data source
  rtf_figure(fig_width = 6, fig_height = 3.6) %>% # Set proportional figure size to the original PNG figure size
  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
  write_rtf(file = "tlf/fig_km2.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/fig_km2.pdf")
```

### Plot with `visR`

Equivalently, the `visR` package can be also used to plot the KM curve fitting CDISC ADaM standard. In the following we will illustrate how to create KM plot in RTF format using `visR`. Firstly, similar to the previous sections, create a png file.

```{r}
# Estimate KM data point
surv_mod <-
  visR::estimate_KM(data = adtte_ttde,
                    strata = "TRTP",
                    AVAL = "AVAL_m")
# save plot with white background and Times New Roman font
ggplot2::theme_set(theme_bw() + theme(text = element_text(family = 'serif')))

png("tlf/fig_km3.png",
    width = 3000,
    height = 2000,
    res = 400)
KM <- visR::visr(
  surv_mod,
  y_label = "Survival Probability\n",
  x_label = "Time in Months",
  x_ticks = 0:6,
  y_ticks = seq(0, 1, 0.1),
  legend_position = "bottom"
) %>%
  add_CNSR()

KM %>%
  visR::add_risktable(group = "statlist")  # seemingly current visR::add_risktable() does not support different fonts, will open an issue in the `visR` Github

Then create RTF file

```{r}
# Create RTF figure
rtf_read_png("tlf/fig_km3.png") %>% # Read the PNG file from the file path
  rtf_title(
    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
    "All Participants"
  ) %>% # Add title or subtitle
  rtf_footnote("footnote") %>% # Add footnote
  rtf_source("[datasource: adam-adtte]") %>% # Add data source
  rtf_figure(fig_width = 6, fig_height = 4) %>% # Set proportional figure size to the original PNG figure size
  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
  write_rtf(file = "tlf/fig_km3.rtf")
```

Alternatively, we can also directly generate PDF file for KM plot with title and footnote (Ref: https://github.com/RConsortium/submissions-pilot1/blob/main/vignettes/tlf-kmplot.Rmd).

```{r}
# save plot as PDF file
pdf.options(reset = TRUE, onefile = FALSE)
pdf("tlf/fig_km3.pdf", width = 12, height = 6)

KM <- visR::visr(
  surv_mod,
  y_label = "Survival Probability\n",
  x_label = "Time in Months",
  x_ticks = 0:6,
  y_ticks = seq(0, 1, 0.1)
) %>%
  add_CNSR()

KM1 <- KM %>%
  visR::add_risktable(group = "statlist")  # seemingly current visR::add_risktable() does not support different fonts, will open an issue in the `visR` Github

title <- cowplot::ggdraw() +
  cowplot::draw_label(
    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group\nAll Participants",
    fontfamily = "serif",
    size = 12
  )
caption <- cowplot::ggdraw() +
  cowplot::draw_label(
    paste0(
      "\nProgram: tlf-efficacy-km.Rmd [",
      Sys.time(),
      "]\n[Datasource: adam-adtte]"
    ),
    fontfamily = "serif",
    size = 12
  )

KM2 <- cowplot::plot_grid(title,
                          KM1,
                          caption,
                          ncol = 1,
                          rel_heights = c(0.1, 0.8, 0.1)) #%>%
# visR::apply_theme(theme_times)
print(KM2)

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/fig_km3.pdf")
```

In conclusion, the procedure to generate the above efficacy K-M plot is summarized as follows.

- Step 1: Read the data `adtte` into R.
- Step 2: Define the analysis-ready dataset. In this example, we define the analysis dataset for the TTDE endpoint `adtte_ttde`.
- Step 3: Create the curve's data points using the `Surv()` formula directly 
or fitted survival models beforehand. Then plot and customize the K-M curve with risk table using `survival`, `survminer` or `visR` package.
- Step 4: Save the output as a PNG file, then format it into an RTF table. (The output format is based on user's preference)

>>>>

```{r}
library(haven) # Read SAS data
library(dplyr) # Manipulate data
library(r2rtf) # Reporting in RTF format
library(survival) # Fit survival model
```

In this chapter, we illustrate how to create a simplified Kaplan-Meier plot in a study.
For the survival analysis in efficacy, time to dermatologic event (TTDE) will be analyzed.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/fig_km.pdf")
```

::: {.rmdnote}
There are R packages that can create more informative Kaplan-Meier plot such as `visR` and `survminer`.
Interested reader can find examples in their package website. 
:::

@@

The `survival` package is used to obtain the K-M estimator.

@@

Using code below, we save the simplified KM plot into a `.png` file.

```{r}
# Save as a PNG file
png(
  file = "tlf/fig_km.png",
  width = 3000,
  height = 2000,
  res = 300
)

plot(
  fit,
  xlab = "Time in Months",
  ylab = "Survival probability",
  lwd = 2,
  mark.time = T,
  col = c(2, 3, 4),
  lty = c(1, 2, 3)
)

@@

Now, we can use the `r2rtf` package to create a formatted RTF figure. 
More details can be found in [r2rtf website](https://merck.github.io/r2rtf/articles/example-figure.html).

```{r}
# Create RTF figure
rtf_read_figure("tlf/fig_km.png") %>% # Read the PNG file from the file path
  rtf_title(
    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
    "All Participants"
  ) %>% # Add title or subtitle
  rtf_footnote("footnote") %>% # Add footnote
  rtf_source("[datasource: adam-adtte]") %>% # Add data source
  rtf_figure(fig_width = 6, fig_height = 4) %>% # Set proportional figure size to the original PNG figure size
  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
  write_rtf(file = "tlf/fig_km.rtf")
```

```{r, include=FALSE}
rtf2pdf("tlf/tlf_eff.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/fig_km.pdf")
```

In conclusion, the steps to create a K-M plot are as follows.

- Step 1: Read the data `adtte` into R.
- Step 2: Define the analysis-ready dataset. In this example, we define the analysis dataset for the TTDE endpoint `adtte_ttde`.
- Step 3: Save figures into `png` files based on required analysis specification.
- Step 4: Create RTF table using `r2rtf` package.
